The polynomial representation of the type $A_{n - 1}$ rational Cherednik algebra in characteristic $p \mid n$

### 2015-05-26 formalwriteup.tex
line edits
Line edits

As described in \cite[Section 2.5]{BC1}, there is a contravariant form 
\[
\beta_c: \Sym(\h^*) \otimes \Sym(\h) \to k
\]
defined by setting $\beta_c(1, 1) = 1$ and imposing for all $x \in \h^*, y \in \h, f \in \Sym(\h^*), g \in \Sym(\h)$ that
\[
\beta_c(xf,g)=\beta_c(f,D_x(g)) \qquad \text{ and } \qquad \beta_c(f,yg) = \beta_c(D_y(f),g).
\]
Then the polynomial representation $\Sym(\h^*)$ has unique maximal graded proper submodule $J_c = \ker(\beta_c)$.  By the definition of $\beta_c$, $J_c$ contains all homogeneous vectors $f \in A$ killed by all Dunkl operators $D_y$.  Such $f$ are known as \textit{singular vectors}.  The quotient $L = A/J_c$ is an irreducible representation of $\HH_{1,c}(\h)$.  It inherits a $\ZZ_{\geq 0}$-grading from $A$, and for $L_j$ the degree $j$ subspace of $L$, we may define its Hilbert series as
\[
h_L(t) = \sum_{j \geq 0} \dim L_j t^j.
\]

@@

Our study is motivated by previous work on the representation theory of the type $A$ rational Cherednik algebra in both characteristic $0$ and $p$.  The type $A$ non-modular case $p \gg n$ was studied in \cite{BFG}, and some properties of the maximal proper graded submodule of the polynomial representation were given in both modular and non-modular cases in \cite{BC1}.  In the modular case $p \mid n$, for $p = 2$, the polynomial representation associated to the $n$-dimensional permutation representation was studied.
\begin{theorem}[{\cite[Theorem 5.1]{L}}] \label{thm:lian}
The irreducible quotient of the polynomial representation associated to the $n$-dimensional reflection representation is a complete intersection with Hilbert series
\[
h(t) = (1 + t)^n (1 + t^2).
\]
The corresponding maximal proper graded submodule is generated by $n - 1$ elements of degree $2$ and one element of degree $4$. 
\end{theorem} 
It was further conjectured by Lian in \cite[Conjecture 5.2]{L} that for all $p$ the corresponding irreducible is a complete intersection with $J_c$ having $n - 1$ generators in degree $p$ and a single generator in degree $p^2$.  Our results are consistent with the restriction of Lian's conjecture to the case when $\h$ is restricted to the $(n - 1)$-dimensional quotient.  It would be interesting to extend our work to prove Lian's conjecture in full.  For general $p \mid n$, a subideal of the maximal proper graded submodule was computed in \cite[Proposition 6.1]{DS}.

In characteristic $0$, our results parallel the explicit decomposition of the polynomial representation of the type $A$ rational Cherednik algebra given in \cite{BEG, CE}.  There, the polynomial representation is irreducible unless $c = \frac{r}{n}$ for some integer $r$, and an explicit set of generators of the maximal proper graded submodule is known.
\begin{proposition}[{\cite[Proposition 3.1]{CE}}] \label{prop:ce}
If $\char(k) = 0$ and $c = \frac{r}{n}$, the maximal proper graded submodule $J_c \subset A$ of the polynomial representation $A$ of $\HH_{1,c}(\h)$ is generated by
\[
\Res_\infty\left[\frac{dz}{z-x_j} \prod_{i=1}^{n} (z-x_i)^c\right] \text{ for $j=1,\dots,n-1$}.
\]
\end{proposition}
We interpret the characteristic $p$ analogue of Proposition \ref{prop:ce} to mean that if $r = p$ and $p \mid n$, then since $p/n$ is an indeterminacy in characteristic $p$ (because it is equivalent to $0/0$), $c=p/n$ in characteristic $0$ in some sense corresponds to $c$ generic in characteristic $p$.

While this substitution is of course invalid, Proposition \ref{prop:ce} may be interpreted as a statement about certain formal power series.  By using a power series version of this construction of generators which makes sense in characteristic $p$, we are able to mimic the arguments of \cite{BEG, CE} to show that they cut out a complete intersection and generate the entire ideal.  We believe that the philosophy of taking characteristic $p$ analogues of characteristic $0$ results for the rational Cherednik algebra should apply more generally and hope to explore this further in future work.

@@

Consider the ideal $I = \langle f_1, \ldots, f_{n - 1} \rangle \subset A$ generated by the $f_i$.  In this section, we will show that $A/I$ is a complete intersection, for which we require some preparatory computations.

@@

We now show that $A/I$ is a complete intersection.  Our technique is a translation of the idea of the proof of \cite[Theorem 3.2]{CE} to the formal power series context.

\begin{proposition}\label{prop:ci}
For generic $c$, the quotient $A/I$ is a complete intersection. 
\end{proposition}
\begin{proof}
It suffices to show that if $x_1,\dots,x_{n} \in k$ satisfy $f_i(x_1,\dots,x_{n})=0$ for all $i$, then $x_1=\dots=x_{n}=0$.  Suppose that the $x_i$ take the distinct values $\{s_1,\dots,s_{r}\}$, where $s_i$ occurs with multiplicity $m_i>0$ so that $g(z)=\prod_i (1-s_iz)^{m_i}$.  In these terms, we have that 
\[
[z^p]\frac{1}{1-s_iz}F(z)= 0
\]
for $i = 1, \ldots, r $ and wish to show that $r=1$ and $s_1=0$.

@@

\begin{theorem}\label{thm:main}
For generic $c$, $f_1, \ldots, f_{n-1}$ are linearly independent and generate the maximal proper graded submodule $J_c$ of the polynomial representation for $\HH_{1, c}(\h)$.  The irreducible quotient $L = A/J_c$ is a complete intersection with Hilbert series 
\[
h_L(t) = \left(\frac{1-t^p}{1-t}\right)^{n-1}.
\]
\end{theorem}
\begin{proof}
By \cite[Proposition 3.4]{BC1}, the Hilbert series of $L$ is 
\[
h_L(t) = \left(\frac{1-t^p}{1-t}\right)^{n-1}h(t^p)
\]
for a polynomial $h(t)$ with nonnegative integer coefficients.  On the other hand, by Propositions~\ref{prop:linind} and \ref{prop:ci}, $A/I$ is a complete intersection with $n-1$ linearly independent degree $p$ generators $f_i$. Its Hilbert series is
\[
h_{A/I}(t)=\left(\frac{1-t^p}{1-t}\right)^{n-1}.
\]
By Proposition~\ref{prop:ann}, the generators $f_i$ of $I$ are singular vectors, so $I \subseteq J_c$, implying that $h_{A/I}(t) \ge h_{A/J_c}(t)$ coefficient-wise.  We conclude that $h(t) = 1$, hence $h_{A/I}(t)=h_{A/J_c}(t)$ and $I=J_c$, completing the proof.
\end{proof}

>>>>

As described in \cite[Section 2.5]{BC1}, there is a contravariant form 
\[
\beta_c: \Sym(\h^*) \otimes \Sym(\h) \to k
\]
defined by setting $\beta_c(1, 1) = 1$ and imposing for all $x \in \h^*, y \in \h, f \in \Sym(\h^*), g \in \Sym(\h)$ that
\[
\beta_c(xf,g)=\beta_c(f,D_x(g)) \qquad \text{ and } \qquad \beta_c(f,yg) = \beta_c(D_y(f),g).
\]
Then the polynomial representation $\Sym(\h^*)$ has unique maximal graded proper submodule $J_c = \ker(\beta_c)$.  By the definition of $\beta_c$, $J_c$ contains the ideal generated by all homogeneous vectors $f \in A$ killed by all Dunkl operators $D_y$.  Such $f$ are known as \textit{singular vectors}.  The quotient $L = A/J_c$ is an irreducible representation of $\HH_{1,c}(\h)$.  It inherits a $\ZZ_{\geq 0}$-grading from $A$, and for $L_j$ the degree $j$ subspace of $L$, we may define its Hilbert series as
\[
h_L(t) = \sum_{j \geq 0} \dim L_j t^j.
\]

@@

Our study is motivated by previous work on the representation theory of the type $A$ rational Cherednik algebra in both characteristic $0$ and $p$.  The type $A$ non-modular case $p \gg n$ was studied in \cite{BFG}, and some properties of the maximal proper graded submodule of the polynomial representation were given in both modular and non-modular cases in \cite{BC1}.  In the modular case $p \mid n$, for $p = 2$, the polynomial representation associated to the $n$-dimensional permutation representation was studied.
\begin{theorem}[{\cite[Theorem 5.1]{L}}] \label{thm:lian}
The irreducible quotient of the polynomial representation associated to the $n$-dimensional reflection representation is a complete intersection with Hilbert series
\[
h(t) = (1 + t)^n (1 + t^2).
\]
The corresponding maximal proper graded submodule is generated by $n - 1$ elements of degree $2$ and one element of degree $4$. 
\end{theorem} 
It was further conjectured by Lian in \cite[Conjecture 5.2]{L} that for all $p$ the corresponding irreducible is a complete intersection with $J_c$ having $n - 1$ generators in degree $p$ and a single generator in degree $p^2$.  Our results are consistent with the restriction of Lian's conjecture to the case when $\h$ is the $(n - 1)$-dimensional quotient.  It would be interesting to extend our work to prove Lian's conjecture in full.  For general $p \mid n$, a submodule of the maximal proper graded submodule was computed in \cite[Proposition 6.1]{DS}.

In characteristic $0$, our results parallel the explicit decomposition of the polynomial representation of the type $A$ rational Cherednik algebra given in \cite{BEG, CE}.  There, the polynomial representation is irreducible unless $c = \frac{r}{n}$ for some integer $r$, and an explicit set of generators of the maximal proper graded submodule is known.
\begin{proposition}[{\cite[Proposition 3.1]{CE}}] \label{prop:ce}
If $\char(k) = 0$ and $c = \frac{r}{n}$, the maximal proper graded submodule $J_c \subset A$ of the polynomial representation $A$ of $\HH_{1,c}(\h)$ is generated by
\[
\Res_\infty\left[\frac{dz}{z-x_j} \prod_{i=1}^{n} (z-x_i)^c\right] \text{ for $j=1,\dots,n-1$}.
\]
\end{proposition}
We interpret the characteristic $p$ analogue of Proposition \ref{prop:ce} to mean that if $r = p$ and $p \mid n$, then since $p/n$ is equivalent to $0/0$ and thus an indeterminacy in characteristic $p$, taking $c=p/n$ in characteristic $0$ should correspond to taking $c$ generic in characteristic $p$.  While this substitution is of course invalid, Proposition \ref{prop:ce} may be interpreted as a statement about certain formal power series.  By using a power series version of this construction of generators which makes sense in characteristic $p$, we are able to mimic the arguments of \cite{BEG, CE} to show that they cut out a complete intersection and generate the entire ideal.  We believe that the philosophy of taking characteristic $p$ analogues of characteristic $0$ results for the rational Cherednik algebra should apply more generally and hope to explore this further in future work.

@@

Consider the ideal $I_c = \langle f_1, \ldots, f_{n - 1} \rangle \subset A$ generated by the $f_i$.  In this section, we will show that $A/I_c$ is a complete intersection, for which we require some preparatory computations.

@@

We now show that $A/I_c$ is a complete intersection.  Our technique is a translation of the idea of the proof of \cite[Theorem 3.2]{CE} to the formal power series context.

\begin{proposition}\label{prop:ci}
For generic $c$, the quotient $A/I_c$ is a complete intersection. 
\end{proposition}
\begin{proof}
It suffices to show that if $x_1,\dots,x_{n} \in k$ satisfy $f_i(x_1,\dots,x_{n})=0$ for all $i$, then $x_1=\dots=x_{n}=0$.  Suppose that the $x_i$ take the distinct values $\{s_1,\dots,s_{r}\}$, where $s_i$ occurs with multiplicity $m_i>0$ so that $g(z)=\prod_i (1-s_iz)^{m_i}$.  In these terms, we have that 
\[
[z^p]\frac{1}{1-s_iz}F(z)= 0
\]
for $i = 1, \ldots, r $ and wish to show that $r=1$ and $s_1=0$.

@@

\begin{theorem}\label{thm:main}
For generic $c$, $f_1, \ldots, f_{n-1}$ are linearly independent and generate the maximal proper graded submodule $J_c$ of the polynomial representation for $\HH_{1, c}(\h)$.  The irreducible quotient $L = A/J_c$ is a complete intersection with Hilbert series 
\[
h_L(t) = \left(\frac{1-t^p}{1-t}\right)^{n-1}.
\]
\end{theorem}
\begin{proof}
By \cite[Proposition 3.4]{BC1}, the Hilbert series of $L$ is 
\[
h_L(t) = \left(\frac{1-t^p}{1-t}\right)^{n-1}h(t^p)
\]
for a polynomial $h(t)$ with nonnegative integer coefficients.  On the other hand, by Propositions~\ref{prop:linind} and \ref{prop:ci}, $A/I_c$ is a complete intersection with $n-1$ linearly independent degree $p$ generators $f_i$. Its Hilbert series is
\[
h_{A/I_c}(t)=\left(\frac{1-t^p}{1-t}\right)^{n-1}.
\]
By Proposition~\ref{prop:ann}, the generators $f_i$ of $I_c$ are singular vectors, so $I_c \subseteq J_c$, implying that $h_{A/I_c}(t) \ge h_{A/J_c}(t)$ coefficient-wise.  We conclude that $h(t) = 1$, hence $h_{A/I_c}(t)=h_{A/J_c}(t)$ and $I_c=J_c$, completing the proof.
\end{proof}
